For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-04 paper/paper.tex
identifiers^S/^D (Con,Ty,..) in mathsf

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no.\ ED18-1-2019-0030 (Application-specific highly reliable IT
solutions), by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no.\ BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}

@@

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,A\ra Con^D\,(\mathsf{ext}\,\GG\,A) \\
  & U^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
  & El^D && : (\GG^D:Con^D\,\GG)\ra\,Ty^D\,({ext}^D\,\GG^D\,(U^D\,\GG^D))\,(\mathsf{El}\,\GG)
%  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
\end{alignat*}
Above we used implicit quantifications for $\GG:\Con$ and
$A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\GG$ before its explicit parameter of type
$Con^D\,\GG$.

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\GG:\Con)\ra Con^D\,\GG \\
  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra Ty^D\,(\mathsf{elimCon}\,\GG)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = {ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = U^D\,(\mathsf{elimCon}\,\GG) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = El^D\,(\mathsf{elimCon}\,\GG)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and it is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} presented a reduction in
extensional type theory, however, this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak, e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.

>>>>

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no.\ ED18-1-2019-0030 (Application-specific highly reliable IT
solutions), by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no.\ BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@gmail.com}{https://orcid.org/0000-0002-9299-641X}{this author was supported by the CoqHoTT ERC Grant 637339.}

@@

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & \Con^D && : \Con\ra\Set \\
  & \Ty^D && : \Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
  & \mathsf{ext}^D && : (\GG^D:\Con^D\,\GG)\ra \Ty^D\,\GG^D\,A\ra \Con^D\,(\mathsf{ext}\,\GG\,A) \\
  & \U^D && : (\GG^D:\Con^D\,\GG)\ra \Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
  & \El^D && : (\GG^D:\Con^D\,\GG)\ra\,\Ty^D\,(\mathsf{ext}^D\,\GG^D\,(\U^D\,\GG^D))\,(\mathsf{El}\,\GG)
%  & Pi^D && : (\GG^D:\Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
\end{alignat*}
Above we used implicit quantifications for $\GG:\Con$ and
$A:\Ty\,\GG$ to ease readability, e.g.\ $\Ty^D$ has an implicit
parameter $\GG$ before its explicit parameter of type
$\Con^D\,\GG$.

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\GG:\Con)\ra \Con^D\,\GG \\
  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra \Ty^D\,(\mathsf{elimCon}\,\GG)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = \U^D\,(\mathsf{elimCon}\,\GG) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = \El^D\,(\mathsf{elimCon}\,\GG)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and it is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} presented a reduction in
extensional type theory, however, this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & \Con^S && : \Con\ra\Set \\
  & \Ty^S && : \Ty\,\GG\ra\Set \\
  & \mathsf{empty}^S && : \Con^S\,\mathsf{empty} \\
  & \mathsf{ext}^S && : \Con^S\,\GG\ra \Ty^S\,\,A\ra \Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & \U^S && : \Con^S\,\GG\ra \Ty^S\,(\mathsf{U}\,\GG) \\
  & \El^S && : \Con^S\,\GG\ra\,\Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra \Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra \Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak, e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.
